Refunctionalization of Abstract Abstract Machines (Functional Pearl)

### 2018-06-05 paper/paper.tex
revise the nd/choices operators

\textbf{Nondeterminism Abstractions.}
To handle the inevitable nondeterminism, we define a generic function @nd@ as follows:
\begin{lstlisting}
def nd[T,S](ts: Set[T], acc: S, f: (T, S, S => S) => S, g: S => S): S = {
  if (ts.isEmpty) g(acc)
  else f(ts.head, acc, (vss: S) => nd(ts.tail, vss, f, g))
}
\end{lstlisting}

Given a set of elements of type @T@, an initial accumulated value of type @S@, a function @f@
that works on element type @T@ and accumulated type @S@, and a function @g@ that
works on and returns a value of type @S@, @nd@ applies @f@ to each element in the
set iteratively, accumulates the value and finally applies @g@ on it.

If we look at @continue@ and @ndcontinue@ in the disentangled AAM, we will find
that @nd@ is an abstraction of those two functions. In function @continue@,
there are two out calls to @ndcontinue@ and @drive_step@, this is the reason why
we have functions @f@ and @g@ in @nd@.

Readers familiar with functional programming may think of @fold@ at this point; indeed,
it is an implementation of @fold@ in continuation style. But here we have two
continuations, one for the next element in the set, and one for the final result.
This is sometimes referred to as an extended continuation-passing style (ECPS)
that has a continuation and a meta-continuation \cite{Danvy:1990:AC:91556.91622}.

@@

\begin{lstlisting}
case Let(x, App(f, ae), e) =>
  val closures = aeval(f, env, store).asInstanceOf[Set[Clos]]
  nd[Clos, Ans](closures, Set[VS](), { case (clos, clsacc, clsnd) =>
    val Clos(Lam(v, body), c_env) = clos
    val baddr = allocBind(v, new_time)
    val new_cenv = c_env + (v -> baddr)
    val new_store = store.update(baddr, aeval(ae, env, store))
    aeval(body, new_cenv, new_store, new_time, (bdvss: Set[VS]) => {
      nd[VS, Ans](bdvss, Set[VS](), { case (vs, bdacc, bdnd) =>
        val VS(vals, time, store) = vs
        val baddr = allocBind(x, time)
        val new_env = env + (x -> baddr)
        val new_store = store.update(baddr, vals)
        aeval(e, new_env, new_store, time, { 
          /* accumulate the values of one path, and call next body expression of f */
          case evss => bdnd(bdacc ++ evss) 
        })
      }, { 
        /* accumulate the values of paths, and call next target closure */
        case evss => clsnd(evss ++ clsacc) 
      })
    })
  },
  continue)
\end{lstlisting}

@@

\begin{lstlisting}
def aeval(e: Expr, env: Env, store: BStore, time: Time, cache: Cache, continue: Cont): Ans = {
  val config = Config(e, env, store, time)
  /* lookups the out cache, uses the values if contains this config */
  if (cache.outContains(config)) cont(Ans(cache.outGet(config), cache))
  else {
    val new_cache = cache.outUpdate(config, cache.inGet(config))
    val new_time = config.tick 
    e match {
      case Let(x, App(f, ae), e) =>
        val closures = atomicEval(f, env, store).asInstanceOf[Set[Clos]]
        nd[Clos, Ans](closures, Ans(Set[VS](), new_cache), { 
          case (Clos(Lam(v, body), c_env), clsans, clsnd) =>
            ...... /* sets vairables, e.g., new_cenv and new_store */
            aeval(body, new_cenv, new_store, new_time, clsans.cache, { case Ans(bdvss, bdcache) =>
              nd[VS, Ans](bdvss, Ans(Set[VS](), bdcache), { 
                case (VS(vals, time, vssstore), bdans, bdnd) =>
                  ...... /* sets variables, e.g., new_env and new_store */
                  aeval(e, new_env, new_store, time, bdans.cache, { 
                    /* updates the out cache after evaluated to the end of one path  */
                    case Ans(evss, ecache) => bdnd(bdans ++ Ans(evss, ecache.outUpdate(config, evss)))
                  })
              }, { case eans => clsnd(eans ++ clsans) })
            })
        }, 
        cont)

      case ae if isAtomic(ae) =>
        val vs = Set(VS(atomicEval(ae, env, store), new_time, store))
        cont(Ans(vs, new_cache.outUpdate(config, vs)))
    }
  }
}
\end{lstlisting}

@@

To address this problem, we reintroduce two operators @nd@ and @ndcps@ for handling 
non-deterministic choices in a way that mimics extended continuation-passing style.

\textbf{Nondeterminism Operators, Again.}
The first operator @nd@ is simplified from the previous one, but specialized with type 
@Ans@ and @Cache@. We may still keep the types generic, but the presentation clearly, 
exposing @Cache@ explictly reveals that cache should be monotonically accumulated to 
the rest of computation. Now @nd@ only takes a function @f@ that applies on type 
@(T, Cache)@ where the value of type @T@ comes from the set, and @nd@ performs recursive 
call on itself over elements in the set @ts@. We also move the accumulation operation 
(@++@) inside of @nd@.

\begin{lstlisting}
def nd[T](ts: Set[T], acc: Ans, k: ((T, Cache)) => Ans): Ans = {
  if (ts.isEmpty) acc
  else nd(ts.tail, acc ++ k(ts.head, acc.cache), k)
}
\end{lstlisting}

The second operator @ndcps@ is implemented with delimited control operator 
@shift@ but simply calls our first operator @nd@ in body.

\begin{lstlisting}
def ndcps[T](ts: Set[T], acc: Ans): (T, Cache) @cps[Ans] = shift { f: (((T, Cache)) => Ans) =>
  nd(ts, acc, f)
}
\end{lstlisting}

The function @ndcps@ takes two arguments: a set of elements with type @T@, 
and an initial accumulated value of type @Ans@.
The @ndcps@ iteratively returns an element from the set along with the latest cache, 
and the return type is also annotated by \verb|@cps[Ans]|. 
This annotation denotes the final returned value of @ndcps@ is type @Ans@.
The @f@ introduced by the @shift@ operator is the delimited continuation from the 
call-site of @ndcps@.

Now we can use these two operators to redefine @aeval@.
We add an annotation \verb|@cps[Ans]| on the return type of the @aeval@ function.
Then, for every nondeterministic choice in the abstract interpreter, we can simply call 
@ndcps@ on the set of possible choices and write the program sequentially as a concrete 
interpreter. @ndcps@ also returns the latest cache to its left-hand side definition.
The cache is still accumulated to the next call to @ndcps@ or @aeval@, 
given the fact that the abstract interpreter should always use the latest cache.

\begin{lstlisting}
def aeval(e: Expr, env: Env, store: BStore, time: Time, cache: Cache): Ans @cps[Ans] = {
  val config = Config(e, env, store, time)
  if (cache.outContains(config)) Ans(cache.outGet(config), cache)
  else {
    val new_time = config.tick
    val new_cache = cache.outUpdate(config, cache.inGet(config))
    e match {
      case Let(x, App(f, ae), e) =>
        val closures = atomicEval(f, env, store).asInstanceOf[Set[Clos]]
        val (Clos(Lam(v, body), c_env), clscache) = ndcps[Clos](closures, Ans(Set[VS](), new_cache))
        val vbaddr = allocBind(v, new_time)
        val new_cenv = c_env + (v -> vbaddr)
        val new_store = store.update(vbaddr, atomicEval(ae, env, store))
        /* evaluates the function application App(f, ae) */
        val Ans(bdvss, bdcache) = aeval(body, new_cenv, new_store, new_time, clscache)
        val (VS(vals, time, vsstore), vscache) = ndcps[VS](bdvss, Ans(Set[VS](), bdcache))
        val baddr = allocBind(x, time)
        val new_env = env + (x -> baddr)
        val new_vsstore = vsstore.update(baddr, vals)
        /* evaluates the body expression from the let */
        val Ans(evss, ecache) = aeval(e, new_env, new_vsstore, time, vscache)
        Ans(evss, ecache.outUpdate(config, evss))

@@

The function @analyze@ is also changed by adding a @reset@ operator around the function 
call to @aeval@. The afterwards updating on cache becomes sequential.

>>>>

\textbf{Nondeterminism Abstractions.}
To handle the inevitable nondeterminism, we define a generic function @nd@ as follows:

\begin{lstlisting}
def nd[T](ts: Iterable[T], acc: Ans, k: (T, Ans, Ans => Ans) => Ans, m: Ans => Ans): Ans = {
  if (ts.isEmpty) m(acc)
  else k(ts.head, acc, (ans: Ans) => nd(ts.tail, ans, k, m))
}
\end{lstlisting}

Given a collection of elements of type @T@, an initial accumlated value of type 
@Ans@, a function @k@ that works on element's type @T@ and accumulated value's 
type @Ans@ (note that @k@ also takes a continuation of type @Ans=>Ans@), 
and a function @m@ that works on and returns a value of type @Ans@, 
@nd@ chooses an element in the collection and applies @k@ to it, accumulates 
the value and finally applies @m@ on the accumulated value when there is no element 
in the collection.

If we look at @continue@ and @ndcontinue@ in the disentangled AAM, we will find
that @nd@ is an abstraction of those two functions. In function @continue@,
there are two out calls to @ndcontinue@ and @drive_step@, this is the reason why
we have functions @k@ and @m@ in @nd@.

Readers familiar with functional programming may think of @fold@ at this point; indeed,
it is an implementation of @fold@ in continuation style. But here we have two
continuations, one for the next element in the collection, and one for the final result.
This is sometimes referred to as an extended continuation-passing style (ECPS)
that function @k@ is a normal continuation and @m@ is a meta-continuation 
\cite{Danvy:1990:AC:91556.91622}.

@@

\begin{lstlisting}
case Let(x, App(f, ae), e) =>
  val closures = atomicEval(f, env, store)
  nd[Storable](closures, Set[VS](), { case (clos, clsacc, clsnd) =>
    val Clos(Lam(v, body), c_env) = clos
    val baddr = allocBind(v, new_time)
    val new_cenv = c_env + (v -> baddr)
    val new_store = store.update(baddr, atomicEval(ae, env, store))
    aeval(body, new_cenv, new_store, new_time, (bdvss: Set[VS]) => {
      nd[VS](bdvss, Set[VS](), { case (vs, bdacc, bdnd) =>
        val VS(vals, time, store) = vs
        val baddr = allocBind(x, time)
        val new_env = env + (x -> baddr)
        val new_store = store.update(baddr, vals)
        /* evaluates and accumluates the values of one path, then call bdnd */
        aeval(e, new_env, new_store, time, { case evss => bdnd(bdacc++evss) }) 
      }, { case evss => clsnd(evss++clsacc) }) /* accumulate the values of (multiple) paths, then call clsnd */
    })
  }, continue)
\end{lstlisting}

@@

\begin{lstlisting}
def aeval(e: Expr, env: Env, store: BStore, time: Time, cache: Cache, continue: Cont): Ans = {
  val config = Config(e, env, store, time)
  /* lookups the out cache, uses the values if contains this config */
  if (cache.outContains(config)) cont(Ans(cache.outGet(config), cache))
  else {
    val new_cache = cache.outUpdate(config, cache.inGet(config))
    val new_time = config.tick 
    e match {
      case Let(x, App(f, ae), e) =>
        val closures = atomicEval(f, env, store)
        nd[Storable](closures, Ans(Set[VS](), new_cache), {
          case (Clos(Lam(v, body), c_env), clsans, clsnd) =>
            ...... /* sets vairables, e.g., new_cenv and new_store */
            aeval(body, new_cenv, new_store, new_time, clsans.cache, { case Ans(bdvss, bdcache) =>
              nd[VS](bdvss, Ans(Set[VS](), bdcache), {
                case (VS(vals, time, vssstore), bdans, bdnd) =>
                  ...... /* sets variables, e.g., new_env and new_store */
                  aeval(e, new_env, new_store, time, bdans.cache, {
                    /* updates the out cache after evaluated to the end of one path  */
                    case Ans(evss, ecache) => bdnd(bdans ++ Ans(evss, ecache.outUpdate(config, evss)))
                  })
              }, { case eans => clsnd(eans ++ clsans) })
            })
        },
        continue)

      case ae if isAtomic(ae) =>
        val vs = Set(VS(atomicEval(ae, env, store), new_time, store))
        continue(Ans(vs, new_cache.outUpdate(config, vs)))
    }
  }
}
\end{lstlisting}

@@

To address this problem, we reintroduce two operators @nd@ and @choices@ for handling 
nondeterministic choices in a way that is direct-style but mimics extended 
continuation-passing style.

\textbf{Nondeterminism Operators, Again.}
The first operator @nd@ is adapted from the previous one, but here @nd@ has only one 
continuation @k@. Introducing @Cache@ explicitly as the part of arguments of 
continuation @k@ reveals that cache should be monotonically accumulated to 
the rest of computation. Now the only continuation @k@ applies on type @(T, Cache)@ where
the value of type @T@ comes from the collection, and @nd@ performs recursive invocation
on itself over elements in the collection @ts@. We also move the accumulation operation
(@++@) inside of @nd@.

\begin{lstlisting}
def nd[T](ts: Iterable[T], acc: Ans, k: ((T, Cache)) => Ans): Ans = {
  if (ts.isEmpty) acc
  else nd(ts.tail, acc ++ k(ts.head, acc.cache), k)
}
\end{lstlisting}

The second operator @choices@ is implemented with delimited control operator 
@shift@ but simply calls our first operator @nd@ in body.

\begin{lstlisting}
def choices[T](ts: Iterable[T], cache: Cache): (T, Cache) @cps[Ans] = shift {
  f: (((T, Cache)) => Ans) => nd(ts, Ans(Set[VS](), cache), f)
}
\end{lstlisting}

The function @choices@ takes two arguments: a set of elements with type @T@, 
and an initial accumulated value of type @Ans@.
The @choices@ iteratively returns an element from the collection along with the latest cache, 
and the return type is also annotated by \verb|@cps[Ans]|. 
This annotation denotes the final returned value of @choices@ is type @Ans@.
The @f@ introduced by the @shift@ operator is the delimited continuation from the 
call-site of @choices@.

Now we can use these two operators to redefine @aeval@.
We add an annotation \verb|@cps[Ans]| on the return type of the @aeval@ function.
Then, for every nondeterministic choice in the abstract interpreter, we can simply call 
@choices@ on the set of possible choices and write the program sequentially as concrete 
interpreters, and the subsequent statements and expressions become the delimited continuation
@f@. @choices@ also returns the latest cache to its left-hand side definition.
The cache is still accumulated to the next call to @choices@ or @aeval@, 
given the fact that the abstract interpreter should always use the latest cache.

\begin{lstlisting}
def aeval(e: Expr, env: Env, store: BStore, time: Time, cache: Cache): Ans @cps[Ans] = {
  val config = Config(e, env, store, time)
  if (cache.outContains(config)) Ans(cache.outGet(config), cache)
  else {
    val new_time = config.tick
    val new_cache = cache.outUpdate(config, cache.inGet(config))
    e match {
      case Let(x, App(f, ae), e) =>
        val closures = atomicEval(f, env, store)
        val (Clos(Lam(v, body), c_env), clscache) = choices[Storable](closures, new_cache)
        val vbaddr = allocBind(v, new_time)
        val new_cenv = c_env + (v -> vbaddr)
        val new_store = store.update(vbaddr, atomicEval(ae, env, store))
        /* evaluates the function application App(f, ae) */
        val Ans(bdvss, bdcache) = aeval(body, new_cenv, new_store, new_time, clscache)
        val (VS(vals, time, vsstore), vscache) = choices[VS](bdvss, bdcache)
        val baddr = allocBind(x, time)
        val new_env = env + (x -> baddr)
        val new_vsstore = vsstore.update(baddr, vals)
        /* evaluates the body expression from the let */
        val Ans(evss, ecache) = aeval(e, new_env, new_vsstore, time, vscache)
        Ans(evss, ecache.outUpdate(config, evss))

@@

The function @analyze@ is also changed by adding a @reset@ operator around the function 
call to @aeval@ to set the delimiter. The afterwards updating on cache becomes sequential.
